\documentclass{beamer}

\usepackage{beamerthemesplit}

% How do I get numbered entries in a beamer bibliography
% http://tex.stackexchange.com/a/124271/111537
\setbeamertemplate{bibliography item}{\insertbiblabel}

\title{Intrusion Detection using Outliers in a Cybersecurity Dataset}
\author{Nikolaos Perrakis}
\date{\today}

\begin{document}

\frame{\titlepage}

\section[Outline]{}
\frame{\tableofcontents}

\section{Introduction}
%\subsection{Overview of the Beamer Class}
\frame
{
  %\frametitle{Introduction}

  \begin{itemize}
  \item Anomaly and Outlier Detection
  \item 4\textsuperscript{th} Industrial Revolution
  \item Intrusion Detection Systems
  \end{itemize}
}

\section{Datasets}
\subsection{DARPA/KDD Cup Dataset}
\frame
{
DARPA 1998 and KDD Cup 1999 Intrusion Detection Datasets are the first well known attempts to create a solid IDS dataset. However they have many problems\cite{ids3}:
\begin{itemize}
\item Generation procedure could have been more realistic.
\item Software used is outdated and misrepresentative of current IT landscape.
\item Artefacts of the simulation cause overestimation of efficiency.
\item Inconsistency in labels and attacks used.
\end{itemize}
}

\subsection{ADFA-LD 12 Dataset}
\frame
{
The main dataset we will use during the MSc Project:\\
\center{ is the \textbf{ADFA LD12}\cite{dat2} Dataset}:
\begin{itemize}
\item Updated software with the inclusion of a component with a known vulnurenability.\\
\begin{itemize}
\item  Common architectural service of web server solutions (LAMP):\\
Ubuntu $11.04$, Apache v$2.2.17$,  PHP v$5.3.5$ and MySQL v$14.14$
\item FTP and SSH services were enabled to simulate remote administration
\item Tiki Wiki v$8.1$ - web based collaborative tool. It has a known vulnerability which simulates 0-day exploitable bug.
\end{itemize}

\item Representative of modern attack structure and methodology.

\end{itemize}
}

\begin{frame}{Attacks used in ADFA-LD 12 dataset.}
\begin{table}
\begin{tabular}{|c|c|}
\hline
\textbf{Payload/Effect} & \textbf{Vector} \\ \hline 
Password brute force  & ftp by hydra \\ \hline
Password brute force & ssh by hydra \\ \hline
Add new superuser & Client side poison executable\\ \hline
Java based meterpreter & Tiki Wiki Vulnerability exploit\\ \hline
Linux meterpreter payload & Client sidepoison executable\\ \hline
C100 Webshell & Php remote file inclusion vulnerability \\ \hline
\end{tabular}
%\vspace{5pt}
%\caption{Attacks used in ADFA-LD 12 dataset.}
\label{tab2}
\end{table}
\end{frame}

\begin{frame}{Data Format}
%{
Each data point is a variable length (time) series of kernel system calls!

\begin{table}
\begin{tabular}{|c|c|}
\hline
Subset &  Data points \\ \hline 
Training  & 833 \\ \hline
Validation & 4372 \\ \hline
Attack &  719\\ \hline
\end{tabular}
\vspace{5pt}
\caption{Subsets of ADFA-LD 12 dataset.}
\label{tab3}
\end{table}

%}
\end{frame}




\section{Feature Engineering}



\section{Machine Learning Algorithms}
\frame
{
We will focus on Statistical Based Outlier Detection.
\begin{itemize}
\item Gaussian Mixture Model Anomaly Detection\cite{out2}
\item Support Vector Machine for Novelty Detection\cite{out3}
\item Outlier Rejecting Regression\cite{out6}
\end{itemize}
}

\section{Research Prospects}
\frame{
\begin{itemize}
\item Improve feature engineering or use more custom kernels for SVM methods for better performance.
\item Improve Scalability of algorithms.\\
A good candidate is the \textbf{AWID 2015}\cite{dat3} Dataset.\\ 
10 Gb in size, we can check scalability of our methods.
\end{itemize}
}

\section{Bibliography}
\frame
{
\scriptsize
\bibliographystyle{abbrv}
\begin{thebibliography}{6}
\bibitem{out2} S. Roberts and L. Tarassenko, \emph{A probabilistic resource allocating network for novelty detection}, Neural Computation, vol. 6, no. 2, pp. 270 - 284, 1994.
\bibitem{out3} P. Hayton, B. Sch¨olkopf, L. Tarassenko, and P. Anuzis, \emph{Support vector novelty detection applied to jet engine vibration spectra}, in NIPS, pp. 946–952, 2000.
\bibitem{out6} Y. Gunawardana, S. Fujiwara, A. Takeda, J. Woo, C. Woelk, and M. Niranjan, \emph{Outlier detection at the transcriptomeproteome interface}, Bioinformatics, 2015.

\bibitem{ids3} Mohiuddin Ahmed, Abdun Naser Mahmood, Jiankun Hu \emph{A survey of network anomaly detection techniques}, Journal of Network and Computer Applications. Vol. 60, January 2016, p. 19-31

\bibitem{dat2} Gideon Creech, Jiankun Huy \emph{Generation of a new IDS Test Dataset: Time to Retire the KDD Collection}, 
2013 IEEE Wireless Communications and Networking Conference (WCNC)
\bibitem{dat3} Constantinos Kolias, Georgios Kambourakis, Angelos Stavrou, and Stefanos Gritzalis (2015) \emph{Intrusion Detection in 802.11 Networks: Empirical Evaluation of Threats and a Public Dataset} IEEE Communication Surveys \& Tutorials, Vol. 18, No. 1, 2016
\end{thebibliography}
}
\end{document}
